---
title: "Introduction to quantitative research methods using R"
author: "Matthew Ivory"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```

This week, we will do our first steps in R. Please work through the following handout at your own pace.

Three important things to remember:

1. As you complete the handout, please don’t just read the commands, please type every single one of them. This is really important: Learning to program is like practicing a conversation in a new language. You will improve gradually, but only if you practice.

2. If you're stuck with something, please write down your questions (to share later in class) and try to solve the problem. Please ask your group members for support and, conversely, if another student is stuck, please try to help them out, too. This way, we develop a supportive learning environment that benefits everybody. In addition, get used to the Help pages in RStudio and start finding solutions online (discussion forums, online textbooks, etc.). This is really important, too. You will only really know how to do quantitative research and statistical analyses when you are doing your own research and dealing with your own data. At that point, you need to be sufficiently autonomous to solve problems, otherwise you will end up making very slow progress in your PhD.

3. Finally, if you do not complete the handout in class, please complete the handout at home. This is important as we will assume that you know the material covered in this handout. And again, the more you practice the better, so completing these handouts at home is important.

### References for this handout

Many of the examples and data files from our class come from these excellent textbooks:

* Andrews, M. (2021). Doing data science in R. Sage.
* Crawley, M. J. (2013). The R book. Wiley.
* Fogarty, B. J. (2019). Quantitative social science data with R. Sage.
* Winter, B. (2019). Statistics for linguists. An introduction using R. Routledge.

## Step 1: Using the R console

R is a command-based system. This means: You type commands (such as the ones highlighted below), R translates the commands into machine instructions, which your computer then executes.

You can type the R commands directly into the console. Or, as you will see later, you can also type commands into the script editor and run the sequence of commands as a batch or collection of lines.

In the next section, we will try out writing commands in the Console pane.

Commands are typed at the command prompt >. We then press Return (Mac) / Enter (Windows), and our command is executed. The output of the command, if there is any output, will be displayed on the next line(s).

## Step 2: Using R as a Calculator

Most textbooks recommend familiarizing yourself with R and RStudio by first using R as a calculator. Let’s try this out.

For example, if you type 10 + 10 in the command prompt and press Return (Mac) / Return (Mac) / Enter (Windows), the result will be displayed underneath:

```{r}
#| eval: true

10 + 10
```


Please try out the following commands.

In the task below, just type the commands in the shaded area, then press Return (Mac) / Enter (Windows).

Note: You don’t have the leave spaces around the operators (+, -, *, /, etc.), but the lines are more readable if you do.

So it’s better if you get used to writing 10 + 10 rather than 10+10, even though the output is the same.

The exception are negative values. 

Here, it is recommended not to leave a space between the minus sign - and the value we are negating. So, we would write -3, not - 3, even though it’s the same.

Please try out all of the commands on your computer now.

1. This is how we do addition

```{r}
10 + 10
```

2. Subtraction

```{r}
8 - 2
```

3. Multiplication

```{r}
10 * 14
```

4. Division

```{r}
112/8
```

5. We can also use exponents, i.e. raising a number to the power of another number, e.g., 2^8, as in the example below

```{r}
2 ^ 8
```

6. Square root. This is done via a function called `sqrt()`. We will discuss functions later. For now, just add a numerical value in the parentheses of `sqrt()`

```{r}
sqrt(64)
```

7. You can also combine +, -, \*, /, ^ operators in your commands. By default, the precedence order of operations will be ^ followed by * or /, followed by + or -, just like in a calculator.

```{r}
2+3-4/2
```

or

```{r}
3+9/3*2^8
```


8. But: You can use brackets `()` to overcome the default order to operations: Test the effect of bracketing on the precedence order of operations in the two examples below.

```{r}
#example a
2 + 3 - 4 / 2

#example b
(2 + 3 - 4) / 2
```

or

```{r}
#example a
3 + 9 / 3 * 2 ^ 8

#Example b
(3 + 9) / 3 * 2 ^ 8
```


## Step 3: History

In the Console pane, have you noticed that pressing the up and down arrows does not allow you to go through the different lines (as would happen in a text file, e.g. Word)? Instead, when you’re at the command prompt >, pressing the up and down arrows allows you to move through the history of executed commands. This can save you a lot of time if you want to re- run one of the previous commands.

Go ahead and rerun the last line using this method, but change the last number to a 4 (instead of an 8)

In the Environment, History, etc. pane, you can use the History tab to see your entire command history. If you click on any line in the History tab, it will re-run the command. Again, very helpful as it saves you a lot of typing.
Let’s try this out.

Go into the History pane and find the line where you ran `sqrt(64)` and rerun it

## Step 4: Incomplete commands, and escaping

What happens if you try to execute an incomplete command such as the one below?

9.

```{r}
10 +
```

You will notice there is something missing (another number for the addition). Rather than giving us the result of the addition, R will just display a plus sign `+` in the console instead of the usual `>`. This means that the command is incomplete. And: If you keep hitting Return (Mac) / Enter (Windows), you will just get more plus signs...

You can either complete the command on the next line (try adding a number to complete the addition), or you can just press Esc to exit the command

## Step 5: Variables

The next important step is to learn how to create variables and assign values to variables. In general, the assignment rule is:
    
    name <- expression/value

Expression - An expression is any R code that returns some value. This can be a single number, the result of a calculation, or a complex statistical analysis.

Assignment operator - The <- is called the assignment operator. This assigns everything that is to its right (the expression) to the variable on its left. Rather than typing < and the minus sign, you can also simply press the shortcut Option+- (Mac) or Alt+- (Windows).

Name - The name is simply the name of the variable.

10. run the following

```{r}
x <- 4 * 8
```

It appears that nothing happened; after all, there seems to be no output in the command prompt. However, something did happen after we executed the command `x <- 4 * 8`.

You will see this when you execute the following command.

```{r}
x
```

As you see, the previous command `x <- 4 * 8` assigned the multiplication `4 * 8` to the
variable `x`.

By typing the name of the variable x in the command line and running it, the value of the variable will be displayed, in this case 32 (being the product of 4 * 8).

There is another way to confirm that you created a variable. If you check the Environment tab in the Environment, History, Connections, etc. pane at the top right of your RStudio window, you will now see that new variable listed.

Once you have created a variable, you can use it for other calculations just like you would with any other number

11. 

```{r}
x * 36
```

We can also assign any of these values to new variables.

12.

```{r}
y<-x*28

y
```

## Step 6. Naming your variables

The name has to follow certain naming conventions. The variable name can consist of letters (upper or lower case), numbers, dots, and underscores. However, it must begin with a letter or a dot that is not followed by a number.

13. Which of the following would be okay?

```{r}
#| eval: false
#| error: false

y157 <- 2
x_y_z <- 2
abc_123 <- 2
_x <- 2
.1px <- 2
a_b_c <- 2
x-y-z <- 2
123 <- 2
```


::: {.callout-tip collapse="true"}

these variable names would work:

* y157
* x_y_z
* abc_123
* a_b_c

these will not:

* _x starts with an underscore
* .1px starts with a number
* x-y-z is trying to subtract z from y from x (it thinks you want to run a function)
* 123 is just a number

:::

Even though you can use nonsense sequences such as the ones above, it is good practice to select variable names that are meaningful, short, without dots (use underscore _ instead), and ideally in lowercase characters. For example:

```{r}
age <- 56
income <- 101034
is_married <- TRUE
years_married <- 27
```

## Step 7: Data Types

Data types are variables that refer to collections of values. There are different types of data types (e.g., lists, matrices, arrays), but we will focus on two particularly important ones: vectors and data frames.

### Vectors

Vectors are one-dimensional sequences of values. They are very simple but fundamental data structures, as you will see below when we talk about data frames. For this reason, it’s worth learning about vectors, how to create and manipulate them. Below we will cover numeric and character vectors.

You can create vectors by using the `c()` function. The c stands for combine. All the items within brackets, separated by commas, will be assigned to the variable on the left of the assignment operator.

If you type the following command, you will create a vector with seven elements. This is called a numeric vector as the elements within the brackets are numbers.

14. 

```{r}
#| eval: true

numbers <- c(2, 3, 5, 7, 11, 13, 17)
```

We can now use this vector to perform all kinds of operations. Try out the following, for example.

15.

```{r}
numbers + 1

numbers / 2

numbers^2
```

Note how the operations are applied to each number in the vector.

For any vector (number sequence), we can also refer to individual numbers that form the vector. We do this by means of indexing operations `[]`. For example, to get the first element of the vector, we can type the following. This will display the first element in the vector.

```{r}
numbers[1]
```

You can also extract more than one element from the vector, and even specify the order. For example:

Note the difference in the example above and the one below. Above, we just used `[1]` and that was fine, but because of how indexing works (and it gets more complex when handling different datatypes, such as 2-dimensional tables), we need to use the `c()` function to demonstrate what we are after.

```{r}
numbers[c(4,2,1)]
```

This will retrieve the fourth element in the vector (the number 7), the first element (2) and the second (3).

::: {.callout-note title="Why is it different?" collapse="true"}

What happens if you don't use the `c()` function? Try it out and see what happens.

```{r}
#| eval: true
#| error: true

numbers[4,2,1]
```

You get an error! Now, in this particular situation, the error message isn't the most helpful, but it is important to get familiar with errors (you will be seeing a lot of them in the next 10 weeks!), and note that they are useful. R cannot carry out operations that don't make sense, and as it happens, the above command is telling it to look for the item in three dimensions (imagine a cube made of vectors). When we get errors it is useful to check our code to see what might be wrong. 

:::

Or you can refer to consecutive set of elements in the vector, such as the fourth to the seventh elements. For this, simply type:

```{r}
numbers[4:7]
```

We can also retrieve all elements with the exception of one. For this, we use the minus sign to exclude the vector element that we want to exclude, as in the following example.

```{r}
numbers [-1]
```

Finally, we can also exclude a sequence of elements by adding the minus sign before the c() function. This will exclude from the output all elements that are specified within the brackets.

```{r}

numbers [-c(1, 3, 5, 7)]

```

The numbers vector is a sequence of numbers. We can find out the type of vector by using
the function `class()`. This will confirm that numbers is a numeric vector.

```{r}
class(numbers)
```

::: {.callout-note collapse="true"}

the `class()` command is really useful, and one worth remembering for future troubleshooting. Sometimes commands don't run as you want them to, and checking the class is what you want is a great sanity check. It has saved me on many occasions.

:::

We can also create vectors with elements that are character strings. Here, each element needs to be surrounded by quotation marks (single or double), as in the example below

```{r}

colleges <- c('bowland', 'cartmel', 'county', 'furness', 'fylde', 'graduate', 'grizedale', 'lonsdale', 'pendle')

colleges
```

This type of vector is character, how would you verify this?

::: {.callout-tip collapse="true"}

```{r}
class(colleges)
```

:::

You can index character vectors, just like numeric vectors.

```{r}
colleges[3]
```

But you cannot perform arithmetic functions on character vectors, of course.

```{r}
colleges*2
```

#### Coercing vectors

In vectors, all of the elements must be of the same type. For example, you cannot have a vector that has both numbers and character strings as elements. If you try to create a vector with both numbers and character strings, then some of your elements will coerced into other types.

If you type the following, you will see that the attempt to create a mixed vector with character strings (bowland, cartmel, county, fylde) and numbers (1, 2, 5, 7, 8) converted the numbers into character strings, as evidenced by the quotation marks. You cannot perform calculations on these numbers as R interprets them as text strings.

```{r}
c('bowland', 'cartmel', 'county', 'furness', 'fylde', 1, 2, 5, 7, 8)
```

Finally, you can also combine vectors using thec()function. For example, we can create a new vector called `new_numbers` by combining the original numbers vectors and adding the squares and cubes of numbers.

```{r}
new_numbers <- c(numbers, numbers ^ 2, numbers ^ 3)

new_numbers
```

We have now produced a new vector new_numbers with 21 elements. These don’t fit all in a line and so are wrapped over two lines. The first row displays elements 1 to 12, and the second row begins with element 13. Now you can also see the meaning of the [1] on the output. The [1] is just the index of the first element of the vector show on the corresponding line. [1] refers to the first element in our vector (the number 2), and [13] refers to the 13th element in our vector (169).

#### Naming Vectors

The elements of a vector can be named, too. Each element in our vector can have a distinct label, which can be useful.

```{r}
ages <- c(bob = 27, bill = 34, charles = 76)

ages
```

We can now access the values of the vector by the label or by the index as before.

```{r}
ages['bill']

ages[2]
```

We can also add names to existing vectors by using the names() function. In the following example, we first assign new values to the vector ages. This will delete the previous numbers and labels. We then assign names to this vector using the names() function.

```{r}
ages <- c(23, 54, 8)
names(ages) <- c("michaela", "jane", "jacques")
ages
```

#### Missing Values

Last but not least. Sometimes, we have missing values in our data. In R, missing values are denoted by NA. This is not treated as a character string but as a special symbol.

You can insert a placeholder for a missing value into a numeric or character vector by simply typing NA in the list of elements.

```{r}
a<-c(1,5,7,NA,11,14)
a


b <- c('michaela', 'bill', NA, 'jane')
b
```

### Dataframes

Data frames are probably the most important data structure in R. They are the default form for representing data sets for statistical analyses.

Data frames have a certain number of columns, and each column has the same number of rows. Each column is a vector, and so data frames are essentially collections of equal-length vectors. (This is also why we spent so much time on vectors above...)

There are two ways of creating data frames. We can create a data frame in R by importing a data file, usually in .csv or .xlsx format. (We will discuss how to import files next week.)

Or we can use the `data.frame()` function to create a data frame from scratch, as in the following example.

```{r}
data_df <- data.frame(name = c('bill', 'jane', 'jacques'), age = c(23, 54, 8))

data_df
```

As you can see, we have now created a data frame with two columns (name, age) and three rows (displaying the name and age of each person, Bill, Jane and Jacques). The columns are our variables and the rows are our observations of these variables.

We can refer to specific elements of the data frame by using indices. In the example below, there are two elements within the index [ ], one for the rows, one for the columns. These are separated by a comma [ , ].

For example, to refer to the element that is in the second row, first column, you would type the following.

```{r}
data_df[2 ,1]
```

You can also retrieve multiple elements. One way of doing this is by leaving one of the indices blank.

If you leave the first index blank (e.g., [ , 1]), then you are telling R that you want the information from all the rows. In the example below, you retrieve the information that is in all the rows that are in column 1.

```{r}
data_df[ ,1]
```

In contrast, if you leave the second index blank ([2, ], you will retrieve the information found in the second row across the two columns.

```{r}
data_df[2 , ]
```

We can also be more specific. For example, you can refer to the first and third row of the second column, we would type:

```{r}
data_df[c(1, 3), 2]
```

On the other hand, we can also refer to a column by name. To do this, we need to use the $
notation.

```{r}
data_df$name

data_df$age
```

::: {.callout-note collapse="true"}

Did you notice that RStudio will automatically suggest the variable names after you typed the $? The code completion feature in RStudio makes writing and executing code much easier!

It will do this for pretty much everything that is either a function, a variable, or even an argument (we come to those later), **provided** that you have typed at least three characters (or press Tab to get there faster)

:::

## Step 8: Functions

While data structures hold data in R, functions are used to do things with the data. You have already encountered a few functions above, namely `sqrt()`, `c()`, and `data.frame()`.

Across all R packages and R’s standard library, there are tens of thousands of functions available for you to use. However, most of our analyses in this course will require only a relatively small number of functions.

Below is a general introduction to functions; we will cover functions in more detail as we progress through this course.

Functions tend to have the following structure:

    function(argument 1, argument 2, argument 3, ...)
    
We can think of functions as actions and arguments as the inputs, i.e. something the functions act on. Most functions require at least one argument. If they have more than one argument, these are separated by commas as in the example above.

For example, see what happens when you type the following two commands.

```{r}
sqrt()
sqrt(4)
```

The function sqrt() requires an argument; if you fail to supply an argument you will get an
error message (Error in sqrt()...)

Here are another few functions that are helpful for our analyses.

We can count the number of elements in a given vector by using the length() function.

```{r}
length(new_numbers)
```

We can calculate sum, mean, median, and standard deviation as follows. (More on this in future sessions when we discuss exploratory data analysis.)

```{r}
sum(new_numbers)

mean(new_numbers)

median(new_numbers)

sd(new_numbers)
```

You can also nest functions inside each other, such as:

```{r}
round(sqrt(mean(new_numbers)))
```

Here, we use three functions, each nested within the other. In the example, we first calculated the mean of the vector new_numbers (460.381), then the square root of this value and finally rounded it. We could have done the same calculation in three steps, as in the example below, but nesting the functions in a single command allows us to be more efficient.

Now, **I don't like** nesting functions and neither should you. Next session I will cover this more clearly, and introduce you to a set of functions and stylistic choices that are the gold standard in psychology and have a huge user base (meaning there's plenty of help out there). For now, just be in awe of the complexity of carrying out multiple functions in R.

### Optional arguments

In some cases, functions can also take an additional argument. A good example is the mean() function, which can an additional argument called trim. If we add the trim argument to the mean function, the command will first remove a certain proportion of the extreme values of the vector and then calculate the mean. This is very useful when we are dealing with outliers in our data, for example. (We will talk more about outliers in future sessions.)

In order to trim observations, we need to specify a value between 0 to 0.5. This will trim the highest and the lowest values before calculating the mean. Naturally, a trim value of 0 means you’re not trimming anything, so the value assigned to trim should be greater than 0. For example, a value of 0.1 means you’re trimming the 10% highest and lowest observations, 0.2 means you’re trimming the 20% highest and lowest, and so forth.

```{r}
mean(new_numbers)

mean(new_numbers, trim=0.0)

mean(new_numbers, trim=0.1)

mean(new_numbers, trim=0.2)

mean(new_numbers, trim=0.3)

mean(new_numbers, trim=0.4)

mean(new_numbers, trim=0.5)
```

## Step 9: Help Pages

RStudio has very helpful pages for the available functions. This is useful when you’re not sure if a function requires an argument, or if you’re in doubt about the use of arguments such as trim.

You can access the help page for a given function in different ways.

The most efficient one is by typing a question mark and the function name in the command line. If you now press Return (Mac) / Enter (Windows), this command will also open the help page for the function. Alternatively, you can use thehelp()function in the command line, as below.

```{r}
?mean()

help('mean')
```

Then there is my preferred option (and again, when we start using the script pane, I will make this clear):

You can also go to the search line of the Help tab in the Files, Plots, Packages, Help pane (bottom right of the screen) and type the name of the function there (mean). There is a useful shortcut to search R Help, namely Ctrl+Option+F1 (Mac) and Ctrl+Alt+F1 (Windows).

In each of the cases above, RStudio will open the help page for the function in question in the Help tab.

## Take home task

The following table displays the scores of students in two foreign language exams, one administered at the beginning of term, the other at the end of term.

Create a data frame called language_exams with the information provided in the table, then answer the questions below using R. To save you the headache of tediously typing it all out, you can highlight and paste the code below as-is into your own script.

```{r}
#| echo: true
#| eval: false

language_exams <- data.frame(
  student_id = c('Elin', 'Spencer', 'Crystal', 'Arun', 'Lina', 'Maximilian', 'Leyton', 'Alexandra', 'Valentina', 'Lola', 'Garfield', 'Lucy', 'Shania', 'Arnold', 'Julie', 'Michaela', 'Nicholas'), 
  exam_1 = c(93, 89, 75, 52, 34, 50, 46, 62, 84, 68, 74, 51, 84, 34, 57, 25, 72), 
  exam_2 = c(98, 96, 94, 65, 50, 68, 58, 77, 95, 86, 89, 70, 90, 50, 67, 37, 90))
```


1. What are the mean scores for exam 1 and exam 2?

```{r}
#| echo: false
#| eval: false

mean(language_exams$exam_1) 
mean(language_exams$exam_2)
```


2. What is the difference between the two means?

```{r}
#| echo: false
#| eval: false

mean(language_exams$exam_2) - mean(language_exams$exam_1)
```

3. What are the mean scores for the two exams if you remove extreme values (the top and bottom 20%) from each?

```{r}
#| echo: false
#| eval: false

mean(language_exams$exam_1, trim=0.2) 
mean(language_exams$exam_2, trim=0.2)
```

4. Based on the previous step (with outliers removed): What is the difference between the two means now? Please round the value before reporting the result.

```{r}
#| echo: false
#| eval: false

round(mean(language_exams$exam_2, trim=0.2) - mean(language_exams$exam_1, trim=0.2))
```

5. Can you do steps 3 and 4 in a single command?

```{r}
#| echo: false
#| eval: false

#the long version
mean(language_exams$exam_1, trim=0.2) # Calculate mean scores exam 1 without outliers
mean(language_exams$exam_2, trim=0.2) # Calculate mean scores exam 2 without outliers
77.63636 - 62.81818 # Difference between the trimmed mean exam 2 and trimmed mean exam 1
round(14.81818) #Rounds the value

#the short version
round(mean(language_exams$exam_2, trim=0.2) - mean(language_exams$exam_1, trim=0.2))

```